README
====

```{r echo = F}
library(pmunits)
suppressWarnings(library(tidyverse, quietly = T))
```

# Units API

### setting units

units are an attribute labeled `pmunits` (to not clash with udunits or other units labels).

Units are set via `set_units`

```{r}
set_units(10, "mg/L")
set_units(1:10, "mg/L")
```

and can be used to set units to columns in dplyr.

```{r}
head(Theoph %>% mutate(conc = set_units(conc, "mg/L")))
```

`set_units` expects udunits2 parseable units, so will error

```{r eval=F}
set_units(10, "mg/imnotaunit")

#> Error in set_units(10, "mg/imnotaunit") : units not parseable
```

A list of column-name/units can be provided to set units across a dataframe.

```{r}
theoph_units <- set_units_from_list(Theoph, list(Wt = "kg", Dose = "mg/kg", Time = "hours", conc = "mg/L"))
lapply(theoph_units, attributes)
```

### arithmetic

the common arithmetic `+-*/` functions have been provided, so that unitful columns will be managed.

When adding/subtracting, the resulting units will be those of the left-hand-side of the operator

`x mg/L + y ug/L --> z mg/L` 

```{r}
set_units(10, "mg/L") + set_units(100, "ug/L")
```

To add or subtract, the units must be convertible

```{r eval=F}
set_units(10, "mg/L") + set_units(1, "km/hour")

#> Error in `+.pmunits`(set_units(10, "mg/L"), set_units(1, "km/hour")) : units incompatible
```

However, for multiplication/division, the unit combination is more lazy, passing along until the user
requests a conversion

```{r}
nonsense_unit <- set_units(10, "mg")/set_units(1, "km")
nonsense_unit
```

When a divisor is present, `pmunits` will automatically add parenthesis around the original unit sets to distinguish

```{r}
set_units(10, "mg/kg")*set_units(100, "kg")
```

This is especially important when dealing with more complex units, for example, with molar conversions
the unit set `mg/L/g/mol` will not parse but `(mg/L)/(g/mol)` will properly parse and can be converted


Units may also be multiplied against 'constants' however there is no magic here, if the multiplication is
actually _not_ a constant or one that would affect the units in any way, the user must reset the units to be safe.

```{r}
set_units(10, "mg/kg")*10
```
```{r}
set_units(10, "mg/kg")/10
```

### conversion via `convert`

pmunit vectors may ultimately be converted to the final expected result via `convert`, which takes the form:

`convert(<units_col>, "<units/to/convert>")`

```{r}
convert(set_units(10, "mg/L"), "ug/L")
```

```{r}
is_mg <- set_units(10, "mg/kg")*set_units(100, "kg")
is_mg
```

`convert` can also be used to confirm a combination of units can be properly reduced
```{r}
convert(is_mg, "mg")
```


### Additional conversions

Specific complex conversion have been pre-baked into the library, namely

* molar conversion - `conc_to_molar`

```{r}
conc_to_molar(set_units(10, "mg/L"), set_units(50, "g/mol"))
```

```{r}
conc_to_molar(set_units(10, "mg/L"), set_units(50, "g/mol"), "umol/L")
```


## NOT considered

* what about statistical transformation - log(DV)
